---
title: "Activity 14: Statistical reasoning 6: generalized linear and multilevel models"
subtitle: Feb. 25rd, 2026, Calvin Munson
format: gfm
execute:
  warning: false
editor: source
---

Welcome! This is the sixth statistical reasoning activity. The goals of this activity are to understand how to implement and interpret both generalized linear models and multilevel models. Specifically, you will:

-   Learn to identify the proper distribution to use with data
-   Implement and interpret generalized linear models, including understanding "link functions"
-   Implement and interpret multilevel models by adding "random effects" to models

This is also the final activity of the class! Congratulations on making it!!

------------------------------------------------------------------------

You will submit one output for this activity:

1.  A **PDF** of a rendered Quarto document with all of your R code. Please create a new Quarto document (e.g. don't use this `README.qmd`) and include all of the code that appears in this document, your own code, and **answers to all of the questions** in the "Q#" sections. Submit this PDF through Gradescope.

A reminder: **Please label the code** in your final submission in two ways:

1.  denote your answers to each question using headers that correspond to the question you're answering, and
2.  thoroughly "comment" your code: remember, this means annotating your code directly by typing descriptions of what each line does after a `#`. This will help future you!

------------------------------------------------------------------------

# 1. Generalized linear models

------------------------------------------------------------------------

## 1.1 Introduction

Generalized linear models allow us to use different distributions that reflect the reality of our response variables.

```{r}
library(tidyverse) # For data wrangling
library(brms) # For stats
library(palmerpenguins) # For the data
library(ggeffects) # for plotting model predictions
```

Let's look at the (huge) variety of distribution families that are available to use:

```{r}
?brmsfamily
```

Thankfully, we are only going to use a couple today! 


-------

### Conceptual practice

For Q1.1a and b, consider the following response variables:

1.    Counts of Clarkia flowers in a meadow
2.    Whether or not a female elephant seal gives birth
3.    The percent cover of red algae in the intertidal
4.    Growth of a tree from one year to the next
5.    The spatial area of a forest in square meters

------

### Q1.1a What values can each of the response variables take on?

Describe the type of values each of the variables can take on (e.g. negatives, positives, zeroes, integers, decimals, 0's/1's, etc)

-----

### Q1.1b Choose which distribution best fits each of the response variables

------

### Q1.2 Choose which distribution best fits your final project response variable

Now, 
1.    Write the response variable that you are using for your final project. 
2.    What values can your response variable take on? (e.g. negative, positive, zero, integers, decimals, 0/1, etc)
3.    What distribution fits your response variable?

-------



------------------------------------------------------------------------

## 1.2 DIY

------------------------------------------------------------------------



# 2. Multilevel models

------------------------------------------------------------------------

## 2.1 Introduction

------------------------------------------------------------------------

## 2.2 DIY

------------------------------------------------------------------------


### Render to PDF

When you have finished, remember to pull, stage, commit, and push with GitHub:

-   Pull to check for updates to the remote branch
-   Stage your edits (after saving your document!) by checking the documents you'd like to push
-   Commit your changes with a commit message
-   Push your changes to the remote branch

Then submit the well-labeled PDF on Gradescope. Thanks!
